---
title: "Deep Dive Expression: Results Wrap-up"
author: "Kathleen Durkin"
date: "2025-06-17"
categories: ["E5-coral"]
format:
  html:
    toc: true
execute: 
  eval: TRUE
engine: knitr
bibliography: ../../../references.bib
---

Trying to wrap up the results section for this manuscript!

```{r}
library(dplyr)
```

To-Do List:

-   Confirm mRNA results #s

-   mRNA GO slim top 20 in each species --- think I assigned this to Sam

-   lncRNA results #s -- should go to Zach

-   miRNA results -- fill in counts

-   lncRNA-mRNA interactions. Need to do lncRNA-mRNA PCC on all three species, or see whether Zach has done it

## miRNA-mRNA interactions.

For all three species, summarize binding and PCC

```{r}
Apul_PCC_miRNA_mRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09-Apul-mRNA-miRNA-interactions/Apul-PCC_miRNA_mRNA.csv")
print("# Apul miRNA-mRNA PCC values signficant at p<0.05:")
Apul_PCC_miRNA_mRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

miranda_strict_all_1kb_parsed_apul_updated <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09-Apul-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_apul_updated.txt")
print("# Apul miRNA-3UTR miRanda interactions:")
nrow(miranda_strict_all_1kb_parsed_apul_updated)
print("Average miRanda total bp shared:")
mean(miranda_strict_all_1kb_parsed_apul_updated$V9)
print("SD of miRanda total bp shared")
sd(miranda_strict_all_1kb_parsed_apul_updated$V9)

cat("\n")

Apul_miRanda_5UTR_1kb_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/Apul-miRanda-5UTR_1kb-strict-parsed.txt")
print("# Apul miRNA-5UTR miRanda interactions:")
nrow(Apul_miRanda_5UTR_1kb_strict_parsed)
print("Average miRanda total bp shared:")
mean(Apul_miRanda_5UTR_1kb_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Apul_miRanda_5UTR_1kb_strict_parsed$V9)

cat("\n")

Apul_miRanda_mRNA_full_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/Apul-miRanda-mRNA_full-strict-parsed.txt")
print("# Apul miRNA-CDS miRanda interactions:")
nrow(Apul_miRanda_mRNA_full_strict_parsed)
print("Average miRanda total bp shared:")
mean(Apul_miRanda_mRNA_full_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Apul_miRanda_mRNA_full_strict_parsed$V9)

cat("\n")

Apul_miranda_PCC_miRNA_3UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09-Apul-mRNA-miRNA-interactions/miranda_PCC_miRNA_mRNA.csv")
Apul_miranda_PCC_miRNA_5UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv")
Apul_miranda_PCC_miRNA_CDS <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv")

print("# Apul miRNA-3UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Apul_miranda_PCC_miRNA_3UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Apul miRNA-5UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Apul_miranda_PCC_miRNA_5UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Apul miRNA-CDS miRanda interactions WITH significant PCC (pval < 0.05)")
Apul_miranda_PCC_miRNA_CDS %>% filter(p_value < 0.05) %>% unique() %>% nrow()

```

```{r}

Peve_PCC_miRNA_mRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-PCC_miRNA_mRNA.csv")
print("# Peve miRNA-mRNA PCC values signficant at p<0.05:")
Peve_PCC_miRNA_mRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

miranda_strict_all_1kb_parsed_Peve_updated <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10-Peve-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_peve_updated.txt")
print("# Peve miRNA-3UTR miRanda interactions:")
nrow(miranda_strict_all_1kb_parsed_Peve_updated)
print("Average miRanda total bp shared:")
mean(miranda_strict_all_1kb_parsed_Peve_updated$V9)
print("SD of miRanda total bp shared")
sd(miranda_strict_all_1kb_parsed_Peve_updated$V9)

cat("\n")

Peve_miRanda_5UTR_1kb_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/Peve-miRanda-5UTR_1kb-strict-parsed.txt")
print("# Peve miRNA-5UTR miRanda interactions:")
nrow(Peve_miRanda_5UTR_1kb_strict_parsed)
print("Average miRanda total bp shared:")
mean(Peve_miRanda_5UTR_1kb_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Peve_miRanda_5UTR_1kb_strict_parsed$V9)

cat("\n")

Peve_miRanda_mRNA_full_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/Peve-miRanda-mRNA_full-strict-parsed.txt")
print("# Peve miRNA-CDS miRanda interactions:")
nrow(Peve_miRanda_mRNA_full_strict_parsed)
print("Average miRanda total bp shared:")
mean(Peve_miRanda_mRNA_full_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Peve_miRanda_mRNA_full_strict_parsed$V9)

cat("\n")

Peve_miranda_PCC_miRNA_3UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_miRNA_mRNA.csv")
Peve_miranda_PCC_miRNA_5UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv")
Peve_miranda_PCC_miRNA_CDS <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv")

print("# Peve miRNA-3UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Peve_miranda_PCC_miRNA_3UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Peve miRNA-5UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Peve_miranda_PCC_miRNA_5UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Peve miRNA-CDS miRanda interactions WITH significant PCC (pval < 0.05)")
Peve_miranda_PCC_miRNA_CDS %>% filter(p_value < 0.05) %>% unique() %>% nrow()

```

```{r}
Ptuh_PCC_miRNA_mRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/Ptuh-PCC_miRNA_mRNA.csv")
print("# Ptuh miRNA-mRNA PCC values signficant at p<0.05:")
Ptuh_PCC_miRNA_mRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

miranda_strict_all_1kb_parsed_Ptuh_updated <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/miranda_strict_all_1kb_parsed_ptuh_updated.txt")
print("# Ptuh miRNA-3UTR miRanda interactions:")
nrow(miranda_strict_all_1kb_parsed_Ptuh_updated)
print("Average miRanda total bp shared:")
mean(miranda_strict_all_1kb_parsed_Ptuh_updated$V9)
print("SD of miRanda total bp shared")
sd(miranda_strict_all_1kb_parsed_Ptuh_updated$V9)

cat("\n")

Ptuh_miRanda_5UTR_1kb_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/Ptuh-miRanda-5UTR_1kb-strict-parsed.txt")
print("# Ptuh miRNA-5UTR miRanda interactions:")
nrow(Ptuh_miRanda_5UTR_1kb_strict_parsed)
print("Average miRanda total bp shared:")
mean(Ptuh_miRanda_5UTR_1kb_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Ptuh_miRanda_5UTR_1kb_strict_parsed$V9)

cat("\n")

Ptuh_miRanda_mRNA_full_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/Ptuh-miRanda-mRNA_full-strict-parsed.txt")
print("# Ptuh miRNA-CDS miRanda interactions:")
nrow(Ptuh_miRanda_mRNA_full_strict_parsed)
print("Average miRanda total bp shared:")
mean(Ptuh_miRanda_mRNA_full_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Ptuh_miRanda_mRNA_full_strict_parsed$V9)

cat("\n")

Ptuh_miranda_PCC_miRNA_3UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/Ptuh-miranda_PCC_miRNA_mRNA.csv")
Ptuh_miranda_PCC_miRNA_5UTR <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv")
Ptuh_miranda_PCC_miRNA_CDS <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv")

print("# Ptuh miRNA-3UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Ptuh_miranda_PCC_miRNA_3UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Ptuh miRNA-5UTR miRanda interactions WITH significant PCC (pval < 0.05)")
Ptuh_miranda_PCC_miRNA_5UTR %>% filter(p_value < 0.05) %>% unique() %>% nrow()
print("# Ptuh miRNA-CDS miRanda interactions WITH significant PCC (pval < 0.05)")
Ptuh_miranda_PCC_miRNA_CDS %>% filter(p_value < 0.05) %>% unique() %>% nrow()

```

::: callout-warning
## Warning

Need to check with Jill how the above miRanda results were generated. Points of concern are

1.  Were inputs 3UTR sites only, or all three 3UTR, 5UTR, and CDS regions?

2.  What were the significance parameters? Was there a cutoff for minimum total bp shared?
:::

\^Looks like Jill only used the miRanda settings -e -20 and -strict, while I used -e -20, -sc 100, and -strict. The -sc is a score cutoff. It may not have affected results, since a brief check suggested the energy cutoff (-e -20) was more stringent and implicitly restricted results to scores above 100. However, I should add a manual filter when using Jill's miRanda outputs (miRNA-3UTR) to be safe.

We also have **topGO enrichment of significant miRNA-mRNA interactions** if we want to use them

## miRNA-lncRNA interactions.

For all three species, summarize binding and PCC. overlap among species

```{r}
Apul_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/28-Apul-miRNA-lncRNA-interactions/Apul-PCC_miRNA_lncRNA.csv")
print("# Apul miRNA-lncRNA PCC values signficant at p<0.05:")
Apul_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

Apul_miRanda_lncRNA_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul-miRanda-lncRNA-strict-parsed.txt")
print("# Apul miRNA-lncRNA miRanda interactions:")
nrow(Apul_miRanda_lncRNA_strict_parsed)
print("Average miRanda total bp shared:")
mean(Apul_miRanda_lncRNA_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Apul_miRanda_lncRNA_strict_parsed$V9)

cat("\n")

Apul_lncRNA_to_precursor_blastn <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_precursor_blastn.tab")
colnames(Apul_lncRNA_to_precursor_blastn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
print("# Apul lncRNA that contain full pre-miRNA")
Apul_lncRNA_to_precursor_blastn %>% filter(length >= 90) %>% filter(mismatch == 0) %>% nrow()
print("Note: the above value includes lncRNA that have highly overlapping genomic coordinates, suggesting instances of isoforms of the same lncRNA")

cat("\n")

Apul_miranda_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/28-Apul-miRNA-lncRNA-interactions/miranda_PCC_miRNA_lncRNA.csv")
print("# Apul miRNA-lncRNA miRanda interactions WITH significant PCC (pval < 0.05)")
Apul_miranda_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()


```

```{r}
Peve_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/15-Peve-miRNA-lncRNA-PCC/PCC_miRNA_lncRNA.csv")
print("# Peve miRNA-lncRNA PCC values signficant at p<0.05:")
Peve_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

Peve_miRanda_lncRNA_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/14-Peve-miRNA-lncRNA-BLASTs-miRanda/Peve-miRanda-lncRNA-strict-parsed.txt")
print("# Peve miRNA-lncRNA miRanda interactions:")
nrow(Peve_miRanda_lncRNA_strict_parsed)
print("Average miRanda total bp shared:")
mean(Peve_miRanda_lncRNA_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Peve_miRanda_lncRNA_strict_parsed$V9)

cat("\n")

Peve_lncRNA_to_precursor_blastn <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/14-Peve-miRNA-lncRNA-BLASTs-miRanda/blasts/lncRNA_to_precursor_blastn.tab")
colnames(Peve_lncRNA_to_precursor_blastn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
print("# Peve lncRNA that contain full pre-miRNA")
Peve_lncRNA_to_precursor_blastn %>% filter(length >= 90) %>% filter(mismatch == 0) %>% nrow()
print("Note: the above value includes lncRNA that have highly overlapping genomic coordinates, suggesting instances of isoforms of the same lncRNA")

cat("\n")

Peve_miranda_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/E-Peve/output/15-Peve-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv")
print("# Peve miRNA-lncRNA miRanda interactions WITH significant PCC (pval < 0.05)")
Peve_miranda_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()
```

```{r}
Ptuh_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/15-Ptuh-miRNA-lncRNA-PCC/PCC_miRNA_lncRNA.csv")
print("# Ptuh miRNA-lncRNA PCC values signficant at p<0.05:")
Ptuh_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()

cat("\n")

Ptuh_miRanda_lncRNA_strict_parsed <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/14-Ptuh-miRNA-lncRNA-BLASTs-miRanda/Ptuh-miRanda-lncRNA-strict-parsed.txt")
print("# Ptuh miRNA-lncRNA miRanda interactions:")
nrow(Ptuh_miRanda_lncRNA_strict_parsed)
print("Average miRanda total bp shared:")
mean(Ptuh_miRanda_lncRNA_strict_parsed$V9)
print("SD of miRanda total bp shared")
sd(Ptuh_miRanda_lncRNA_strict_parsed$V9)

cat("\n")

Ptuh_lncRNA_to_precursor_blastn <- read.table("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/14-Ptuh-miRNA-lncRNA-BLASTs-miRanda/blasts/lncRNA_to_precursor_blastn.tab")
colnames(Ptuh_lncRNA_to_precursor_blastn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
print("# Ptuh lncRNA that contain full pre-miRNA")
Ptuh_lncRNA_to_precursor_blastn %>% filter(length >= 90) %>% filter(mismatch == 0) %>% nrow()
print("Note: the above value includes lncRNA that have highly overlapping genomic coordinates, suggesting instances of isoforms of the same lncRNA")

cat("\n")

Ptuh_miranda_PCC_miRNA_lncRNA <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/15-Ptuh-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv")
print("# Ptuh miRNA-lncRNA miRanda interactions WITH significant PCC (pval < 0.05)")
Ptuh_miranda_PCC_miRNA_lncRNA %>% filter(p_value < 0.05) %>% unique() %>% nrow()
```

::: callout-note
## precursors - PCC

Should validate lncRNA acting as miRNA precursors by checking for coexpression
:::

## mRNA-lncRNA PCC

```{r}
Apul_mRNA_lncRNA_PCC <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/F-Ptuh/output/10.1-Ptuh-mRNA-lncRNA-correlation-PCC/Ptuh-PCC_mRNA_lncRNA.csv")
```

## lncRNA-mRNA-miRNA
