---
title: "Deep Dive Expression: Figure edits part 2"
author: "Kathleen Durkin"
date: "2026-04-08"
categories: ["E5-coral"]
format:
  html:
    toc: true
execute: 
  eval: TRUE
engine: knitr
---

Correcting CIRCOES plot generated [here](./2025_12_08_dde_plot_edits.qmd) for DDE manuscript. Old plot shows which targets are orthologus genes (present in more than one species) -- I need the plot to be showing orthologs **that are targeted by miR-100 in more than one species**.

TLDR: Only one orthogroup is targeted by miR-100 in more than one species, targeted in 
*A. pulchra* and *P. evermanni*. This ortholg is not functionally annotated. 

Updated plot showing this conservation is stored [here](https://github.com/shedurkin/Roberts-LabNotebook/tree/master/posts/projects/E5_coral/output/2026_04_08_DDE_plot_edits/circos_plot.png)

Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(circlize)
```

Load edges files and select only miR-100 targets

```{r}
Apul_edges <- read.csv("https://github.com/urol-e5/deep-dive-expression/raw/refs/heads/main/D-Apul/output/33-Apul-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.05_unique.csv", header=TRUE) %>% dplyr::select(-X) %>% filter(source == "apul-mir-100")

Peve_edges <- read.csv("https://github.com/urol-e5/deep-dive-expression/raw/refs/heads/main/E-Peve/output/31-Peve-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.05_unique.csv", header=TRUE) %>% dplyr::select(-X) %>% filter(source == "peve-mir-100")

Ptuh_edges <- read.csv("https://github.com/urol-e5/deep-dive-expression/raw/refs/heads/main/F-Ptuh/output/31-Ptuh-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.05_unique.csv", header=TRUE) %>% dplyr::select(-X) %>% filter(source == "ptuh-mir-100")
```

Read in [Hollie's orthogroup analysis](https://github.com/urol-e5/timeseries_molecular/blob/main/M-multi-species/scripts/33-orthofinder_E5.md) output:

```{r}
h_ortho <- readr::read_tsv("https://github.com/urol-e5/timeseries_molecular/raw/1b3da407f438e9ecf6fbc43b083b4b94897f2f99/M-multi-species/output/33-orthologs/shared_orthologs_all_species.tsv")
```

Add column showing whether a given orthogroup contains proteins from all species or only a subset.

```{r}
# Standardize species names for simplicity
h_ortho <- h_ortho %>%
  mutate(
    simple_species = case_when(
      grepl("Apulchra", Species) ~ "apul",
      grepl("Porites_evermanni", Species) ~ "peve",
      grepl("Pocillopora_meandrina", Species) ~ "ptuh",
      TRUE ~ "other"
    )
  )

# Summarize species presence per orthogroup
h_ortho_summary <- h_ortho %>%
  group_by(Orthogroup) %>%
  summarise(
    type = case_when(
      all(c("apul", "peve", "ptuh") %in% simple_species) ~ "three_way",
      all(c("apul", "peve") %in% simple_species) & !("ptuh" %in% simple_species) ~ "apul_peve",
      all(c("apul", "ptuh") %in% simple_species) & !("peve" %in% simple_species) ~ "apul_ptuh",
      all(c("peve", "ptuh") %in% simple_species) & !("apul" %in% simple_species) ~ "peve_ptuh",
      TRUE ~ "other"
    ),
    .groups = "drop"
  )

# Join back to original
h_ortho <- h_ortho %>%
  left_join(h_ortho_summary, by = "Orthogroup")
```

Huh, looks like Hollie's analysis only outputs orthogroups which contain at least one protein from each species

Read in [Steven's orthogroup analysis](https://github.com/urol-e5/timeseries_molecular/blob/1b3da407f438e9ecf6fbc43b083b4b94897f2f99/M-multi-species/scripts/11.1-orthology.md#L4) output:

```{r}
s_ortho <- read.csv("https://github.com/urol-e5/timeseries_molecular/raw/refs/heads/main/M-multi-species/output/11-orthology-analysis/ortholog_groups.csv", header=TRUE)
```

Annotate each edges file with orthogroup info from both analyses

```{r}
# Annotate each target with orthogroup type
Apul_edges_ortho <- Apul_edges %>%
  rowwise() %>%
  mutate(
    h_ortho_type = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    h_ortho_group = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$Orthogroup[1]
      } else {
        "none"
      }
    },
    s_ortho_type = {
      match_row <- s_ortho %>%
        filter(str_detect(apul, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    s_ortho_group = {
      match_row <- s_ortho %>%
        filter(str_detect(apul, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$group_id[1]
      } else {
        "none"
      }
    }
  ) %>%
  mutate(species = "Apul") %>%
  ungroup()

Peve_edges_ortho <- Peve_edges %>%
  rowwise() %>%
  mutate(
    h_ortho_type = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    h_ortho_group = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$Orthogroup[1]
      } else {
        "none"
      }
    },
    s_ortho_type = {
      match_row <- s_ortho %>%
        filter(str_detect(peve, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    s_ortho_group = {
      match_row <- s_ortho %>%
        filter(str_detect(peve, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$group_id[1]
      } else {
        "none"
      }
    }
  ) %>%
  mutate(species = "Peve") %>%
  ungroup()

Ptuh_edges_ortho <- Ptuh_edges %>%
  rowwise() %>%
  mutate(
    h_ortho_type = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    h_ortho_group = {
      match_row <- h_ortho %>%
        filter(str_detect(Proteins, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$Orthogroup[1]
      } else {
        "none"
      }
    },
    s_ortho_type = {
      match_row <- s_ortho %>%
        filter(str_detect(ptua, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$type[1]
      } else {
        "none"
      }
    },
    s_ortho_group = {
      match_row <- s_ortho %>%
        filter(str_detect(ptua, fixed(target)))
      if(nrow(match_row) > 0) {
        # If multiple matches, pick the first
        match_row$group_id[1]
      } else {
        "none"
      }
    }
  ) %>%
  mutate(species = "Ptuh") %>%
  ungroup()


```

Combine the three species, and identify which orthologs are putatively targeted in multiple species

```{r}
data <- rbind(Apul_edges_ortho, Peve_edges_ortho, Ptuh_edges_ortho)

# Add target_type column to distinguish lncRNA from genes
# Also fix any typos in s_ortho_type (ptua -> ptuh)
data <- data %>%
  mutate(
    target_type = ifelse(grepl("^lncRNA_", target), "lncRNA", "gene"),
    s_ortho_type = case_when(
      s_ortho_type == "apul_ptua" ~ "apul_ptuh",
      s_ortho_type == "peve_ptua" ~ "peve_ptuh",
      TRUE ~ s_ortho_type
    )
  )

# Identify cross-species ortholg targets

get_conserv <- function(data, ortho_col) {
  # Only look at rows with actual orthogroup assignments
  species_map <- data %>%
    filter(.data[[ortho_col]] != "none") %>%
    group_by(across(all_of(ortho_col))) %>%
    summarise(species_set = list(sort(unique(species))), .groups = "drop") %>%
    mutate(conserv = sapply(species_set, function(spp) {
      if (length(spp) >= 3) {
        "three-way"
      } else if (length(spp) == 2) {
        paste(tolower(spp), collapse = "_")
      } else {
        "none"
      }
    }))
  
  setNames(species_map$conserv, species_map[[ortho_col]])
}

h_lookup <- get_conserv(data, "h_ortho_group")
s_lookup <- get_conserv(data, "s_ortho_group")

data <- data %>%
  mutate(
    h_ortho_conserv = ifelse(h_ortho_group == "none", "none", h_lookup[h_ortho_group]),
    s_ortho_conserv = ifelse(s_ortho_group == "none", "none", s_lookup[s_ortho_group])
  )
```

Check

```{r}
data %>% filter(h_ortho_conserv != "none" | s_ortho_conserv != "none") %>% select(target, species, h_ortho_group, h_ortho_conserv, s_ortho_group, s_ortho_conserv)
```

Hmm, so it turns out there's only one orthogroup that is targeted by miR-100 in more than one species -- the orthogroup contains both FUN_035913 in A. pulchra and Peve_00006734 in P. evermanni. On top of that, this single conserved target is also not functionally annotated (does not closely match any previously functionall described sequence -- checked [here](https://github.com/urol-e5/deep-dive-expression/blob/main/D-Apul/output/09.13-Apul-mRNA-miRNA-interactions-FE-pooled/miRNA_pooled_sig_cor_targets_FA.csv)). On the bright side, this conserved targeting shows up in the outputs of both Hollie's and Steven's ortholog analyses, so I don't need to worry about reconciling/picking between divergent results. 

This isn't particularly useful/exciting, so I'm not sure its even worth displaying in the planned plot, but I'll go ahead and do it for posterity.

## Plot

```{r}
# Set margins to allow room for outer labels
par(mar = c(2, 2, 4, 2))

# Define segments for compact version - unconserved targets share same segment
segment_info_v3 <- list(
  # lncRNA segments (left half of circle)
  "lnc_all_3" = list(
    label = "All 3",
    species = c("Apul", "Peve", "Ptuh"),
    ortho_conserv = "three_way",
    target_type = "lncRNA",
    conserved = TRUE
  ),
  "lnc_apul_peve" = list(
    label = "Apul+Peve",
    species = c("Apul", "Peve"),
    ortho_conserv = "apul_peve",
    target_type = "lncRNA",
    conserved = TRUE
  ),
  "lnc_peve_ptuh" = list(
    label = "Peve+Ptuh",
    species = c("Peve", "Ptuh"),
    ortho_conserv = "peve_ptuh",
    target_type = "lncRNA",
    conserved = TRUE
  ),
  "lnc_apul_ptuh" = list(
    label = "Apul+Ptuh",
    species = c("Apul", "Ptuh"),
    ortho_conserv = "apul_ptuh",
    target_type = "lncRNA",
    conserved = TRUE
  ),
  "lnc_unconserved" = list(
    label = "Unconserved",
    species = c("Apul", "Peve", "Ptuh"),
    ortho_conserv = "none",
    target_type = "lncRNA",
    conserved = FALSE
  ),
  # Gene segments (right half of circle)
  "gene_all_3" = list(
    label = "All 3",
    species = c("Apul", "Peve", "Ptuh"),
    ortho_conserv = "three_way",
    target_type = "gene",
    conserved = TRUE
  ),
  "gene_apul_peve" = list(
    label = "Apul+Peve",
    species = c("Apul", "Peve"),
    ortho_conserv = "apul_peve",
    target_type = "gene",
    conserved = TRUE
  ),
  "gene_peve_ptuh" = list(
    label = "Peve+Ptuh",
    species = c("Peve", "Ptuh"),
    ortho_conserv = "peve_ptuh",
    target_type = "gene",
    conserved = TRUE
  ),
  "gene_apul_ptuh" = list(
    label = "Apul+Ptuh",
    species = c("Apul", "Ptuh"),
    ortho_conserv = "apul_ptuh",
    target_type = "gene",
    conserved = TRUE
  ),
  "gene_unconserved" = list(
    label = "Unconserved",
    species = c("Apul", "Peve", "Ptuh"),
    ortho_conserv = "none",
    target_type = "gene",
    conserved = FALSE
  )
)

# Calculate segment sizes for v3 - unconserved segments need max across all species
calc_segment_sizes_v3 <- function(data, segment_info) {
  sizes <- list()

  for (seg_name in names(segment_info)) {
    seg <- segment_info[[seg_name]]

    if (seg$ortho_conserv == "none") {
      # Unconserved: find the max count across all species
      max_n <- 0
      for (sp in seg$species) {
        n <- data %>%
          filter(s_ortho_conserv == "none",
                 species == sp,
                 target_type == seg$target_type) %>%
          select(target) %>% distinct() %>% nrow()
        max_n <- max(max_n, n)
      }
      sizes[[seg_name]] <- max(max_n, 1)
    } else {
      # Conserved: filter by ortho type and target type
      n <- data %>%
        filter(s_ortho_conserv == seg$ortho_conserv,
               target_type == seg$target_type) %>%
        select(target) %>% distinct() %>% nrow()
      sizes[[seg_name]] <- max(n, 1)
    }
  }

  return(sizes)
}

# Create compact circos plot with overlapping unconserved targets
create_circos_plot_v3 <- function(data) {

 # Species order (outer to inner): Peve (most), Apul, Ptuh (fewest)
  species_order <- c("Peve", "Apul", "Ptuh")

  # Calculate segment sizes
  seg_sizes <- calc_segment_sizes_v3(data, segment_info_v3)

  # Define sector order: lncRNA segments first, then gene segments
  lnc_segments <- c("lnc_all_3", "lnc_apul_peve", "lnc_peve_ptuh", "lnc_apul_ptuh", "lnc_unconserved")
  gene_segments <- c("gene_all_3", "gene_apul_peve", "gene_apul_ptuh",  "gene_peve_ptuh", "gene_unconserved")

  # Filter to only segments with data
  lnc_segments <- lnc_segments[sapply(lnc_segments, function(s) {
    seg <- segment_info_v3[[s]]
    if (seg$ortho_conserv == "none") {
      nrow(data %>% filter(s_ortho_conserv == "none", target_type == "lncRNA")) > 0
    } else {
      nrow(data %>% filter(s_ortho_conserv == seg$ortho_conserv, target_type == "lncRNA")) > 0
    }
  })]

  gene_segments <- gene_segments[sapply(gene_segments, function(s) {
    seg <- segment_info_v3[[s]]
    if (seg$ortho_conserv == "none") {
      nrow(data %>% filter(s_ortho_conserv == "none", target_type == "gene")) > 0
    } else {
      nrow(data %>% filter(s_ortho_conserv == seg$ortho_conserv, target_type == "gene")) > 0
    }
  })]

  # All sectors in order
  all_sectors <- c(lnc_segments, gene_segments)

  if (length(all_sectors) == 0) {
    message("No data to plot")
    return(invisible(NULL))
  }

  # Build xlims
  xlims <- list()
  for (seg_name in all_sectors) {
    xlims[[seg_name]] <- c(0, seg_sizes[[seg_name]])
  }

  # Initialize circos
  circos.clear()

  # Calculate gap degrees - larger gaps between lncRNA and gene halves
  n_lnc <- length(lnc_segments)
  n_gene <- length(gene_segments)
  n_total <- n_lnc + n_gene

  # Create gap vector: small gaps within each half, large gap between halves
  gap_degrees <- rep(3, n_total)
  if (n_lnc > 0 && n_gene > 0) {
    gap_degrees[n_lnc] <- 10
    gap_degrees[n_total] <- 20
  }

  circos.par(
    start.degree = 90,
    gap.degree = gap_degrees,
    track.margin = c(0.02, 0.02),
    cell.padding = c(0.02, 0.1, 0.02, 0.1),
    canvas.xlim = c(-1.4, 1.4),
    canvas.ylim = c(-1.4, 1.4)
  )

  circos.initialize(
    factors = factor(all_sectors, levels = all_sectors),
    xlim = do.call(rbind, xlims[all_sectors])
  )

  # Colors
  highlight_col <- rgb(0.6, 0.85, 1, 0.6)  # Light blue for conserved
  no_highlight_col <- rgb(0.95, 0.95, 0.95, 1)  # Light grey for unconserved
  pos_col <- rgb(0.2, 0.7, 0.2, 0.9)  # Green for positive PCC
  neg_col <- rgb(0.9, 0.4, 0.6, 0.9)  # Pink for negative PCC
  border_col <- rgb(0.3, 0.5, 0.7, 1)  # Blue border

  # Create 3 tracks (one per species, outer to inner)
  for (sp_idx in 1:3) {
    sp <- species_order[sp_idx]

    circos.track(
      factors = factor(all_sectors, levels = all_sectors),
      ylim = c(0, 1),
      track.height = 0.15,
      bg.border = "grey60",
      bg.lwd = 1.5,
      panel.fun = function(x, y) {
        sector_name <- CELL_META$sector.index
        seg <- segment_info_v3[[sector_name]]

        # For conserved segments, check if species is in the conservation group
        # For unconserved, all species appear but no highlighting
        if (seg$conserved) {
          sp_in_segment <- sp %in% seg$species
        } else {
          sp_in_segment <- TRUE  # All species appear in unconserved segment
        }

        # Background color
        if (seg$conserved && sp_in_segment) {
          bg_col <- highlight_col
          border <- border_col
        } else if (sp_in_segment) {
          bg_col <- no_highlight_col
          border <- "grey60"
        } else {
          bg_col <- "white"
          border <- "grey80"
        }

        # Draw background rectangle
        circos.rect(
          CELL_META$xlim[1], 0,
          CELL_META$xlim[2], 1,
          col = bg_col,
          border = border,
          lwd = 1.5
        )

        # Plot points
        if (seg$ortho_conserv == "none") {
          # Unconserved: each species plots its own targets
          seg_data <- data %>%
            filter(species == sp,
                   s_ortho_conserv == "none",
                   target_type == seg$target_type) %>%
            distinct(target, .keep_all = TRUE)
        } else if (sp_in_segment) {
          # Conserved: filter by ortho type
          seg_data <- data %>%
            filter(species == sp,
                   s_ortho_conserv == seg$ortho_conserv,
                   target_type == seg$target_type) %>%
            distinct(target, .keep_all = TRUE)
        } else {
          seg_data <- data.frame()
        }

        if (nrow(seg_data) > 0) {
          n_points <- nrow(seg_data)
          if (n_points == 1) {
            x_positions <- CELL_META$xlim[2] / 2
          } else {
            x_positions <- seq(0.5, CELL_META$xlim[2] - 0.5, length.out = n_points)
          }

          for (i in 1:n_points) {
            pt_col <- ifelse(seg_data$PCC_direction[i] > 0, pos_col, neg_col)
            circos.points(
              x_positions[i], 0.5,
              col = pt_col,
              pch = 21,
              bg = pt_col,
              cex = 0.75
            )
          }

          # Add count label above the feature points
          circos.text(
            x = mean(x_positions),
            y = 0.85,
            labels = as.character(n_points),
            facing = "bending.inside",
            niceFacing = TRUE,
            cex = 0.75,
            col = "black",
            font = 2
          )
        }
      }
    )
  }

  # Add segment labels on the outside
  for (seg_name in all_sectors) {
    seg <- segment_info_v3[[seg_name]]

    circos.text(
      x = mean(xlims[[seg_name]]),
      y = 1.8,
      sector.index = seg_name,
      track.index = 1,
      labels = seg$label,
      facing = "bending.outside",
      niceFacing = TRUE,
      cex = 0.7,
      font = 2,
      col = "grey30"
    )
  }

  # Add "lncRNA" and "Genes" labels for each half
  if (n_lnc > 0) {
    mid_lnc <- lnc_segments[ceiling(n_lnc / 2)]
    circos.text(
      x = mean(xlims[[mid_lnc]]),
      y = 2.5,
      sector.index = mid_lnc,
      track.index = 1,
      labels = "lncRNA Targets",
      facing = "bending.outside",
      niceFacing = TRUE,
      cex = 1.0,
      font = 2,
      col = "darkblue"
    )
  }

  if (n_gene > 0) {
    mid_gene <- gene_segments[ceiling(n_gene / 2)]
    circos.text(
      x = mean(xlims[[mid_gene]]),
      y = 2.5,
      sector.index = mid_gene,
      track.index = 1,
      labels = "Gene Targets",
      facing = "bending.outside",
      niceFacing = TRUE,
      cex = 1.0,
      font = 2,
      col = "darkred"
    )
  }

  # Add title and species labels in the center (outer to inner: Peve, Apul, Ptuh)
  text(0, 0.04, "miR-100 Targets", font = 2, cex = 1.1, col = "black")
  text(0, -0.04, "by Conservation", font = 2, cex = 1.1, col = "black")
  text(-0.09, 0.9, "Peve", font = 2, cex = 0.9, col = "grey30")
  text(-0.09, 0.71, "Apul", font = 2, cex = 0.9, col = "grey30")
  text(-0.09, 0.53, "Ptuh", font = 2, cex = 0.9, col = "grey30")

  # Add legend
  legend(
    "bottomright",
    legend = c("Positive PCC", "Negative PCC", "Conserved region"),
    pch = c(21, 21, 22),
    pt.bg = c(pos_col, neg_col, highlight_col),
    col = c(pos_col, neg_col, border_col),
    pt.cex = c(1.5, 1.5, 2.2),
    bty = "n",
    cex = 0.85
  )

  circos.clear()
}

# Create the compact plot
create_circos_plot_v3(data)

# Save to png
png("output/2026_04_08_DDE_plot_edits/s_orthos_circos_plot.png", width = 2000, height = 2000, res = 300)
create_circos_plot_v3(data)
dev.off()

```


